### 2019-05-27 treestats_paper.tex
more words

Theory and computation of single-site statistics of DNA sequences

@@

% summary
In this paper, we present a general framework for defining and efficiently computing
``single-site'' genetic statistics,
i.e., statistics of aligned genome sequence that can be expressed as averages over values computed
separately for each site.
Patterns of genetic polymorphism in a population
are determined by the population's history of Mendelian inheritance,
as summarized by the underlying genealogical trees,
and so it is natural (and helpful) to express these statistics
in terms of the structure of the trees themselves.
This perspective turns out to both suggest efficient algorithms
for computing these statistics on tree sequences,
and allows us to treat single-site genetic statistics in a more general framework
of summaries of tree shape and inheritance.
The methods are implemented in the \tskit\ library.

@@

Suppose we have genotyped a collection of $N$ (homologous) chromosomes
at a sequence of $L$ loci along the genome.
\jk{Maybe use $n$ here? We use it below}
Suppose first that we wish to compute an additive function of the allele frequency spectrum.
In other words, suppose that the frequency of allele $a$ at a given site is $p_i(a)$,
and that the allele frequency spectrum
-- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $f(p)$,
and for some function $h()$ we want to compute $\sum_p f(p) h(p)$.
Since this is a simple average over sites,
we can compute this, summing over the $L$ sites, as
$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L \sum_a h(p_i(a))$.
This definition applies to multiallelic sites by assuming that for each allele
we can only see how many samples carry that allele,
not how the remaining samples are partitioned between one or more other alleles.
\jk{Not clear to me here if we're talking about frequency $0 \leq p \leq 1$ or
`prevalance`, $0 \leq p \leq N$.}
(Note also that this does not differentiate ancestral and derived alleles,
and so is a function of the \emph{folded} allele frequency spectrum.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
occur below that mutation in the tree (and removing those that subsequently mutate).
What we do next is a generalization of this idea.

A more convenient way to parameterize many functions of genotype patterns
uses three ingredients:
\plr{MODIFY and SHORTEN THIS}
\begin{itemize}

    \item[(a)] A list of $k$ vectors of \emph{initial node weights}, $(\iw_1, \ldots, \iw_k)$,
        where $\iw_i(j)$ is the ``initial'' value of the $i^\text{th}$ weight assigned to node $j$.
        \jk{Maybe referring to nodes with variables $u, v, w$ and keeping $i,
        j, k$ for general indexes would help here separate the concepts here?}

    \item[(b)] The \emph{weight} $\nw$ of a node in a tree
        is a $k$-vector of weights is $\nw = (\nw_1, \ldots, \nw_k)$,
        and is equal to the sum of initial weights of all nodes below it in the tree
        (including itself).

    \item[(c)] A \emph{summary function} $f : \R^k \to \R$ so that
        $f(\nw)$ is the ``summary'' statistic we compute \jk

@@

Roughly speaking, we use these to compute a statistic
by first propagating the initial weights up the trees,
and then evaluate the summary function on the resulting vectors of weights
for each allele.

@@

We could calculate for a given tree how many of 
a certain set of ``target'' samples are below each of its edges
by (a) assigning each of those samples weight 1, and other nodes weight zero, then
(b) finding the total weight of all nodes in the subtree below each edge.
This total weight is equal to the number of target samples below that edge,
and it can be calculated efficiently by propagating weights up the tree.
It turns out to be useful to allow more general weights.

\begin{definition}[Subtree weights]
    Given a list of \emph{node weights} $w$, and a tree $T$,
    define the \emph{subtree weight} $x_u$ for each node $u$ to be the sum of all node weights
    of every node below $u$ in the tree (including $u$ itself):
    \begin{align*}
        \nw_u = \sum_{v \,:\, v \le_T u} \iw_v ,
    \end{align*}
    where $v \le_T u$ if $u$ is on the path from $v$ back to the root of $T$.
\end{definition}

@@

The first method of summarizing uses subtree weights
to compute a summary for each variant site in the genome.
To compute the statistic at site $j$,
for each allele $a$ seen at this site,
let $\aw_j(a)$, the weight associated with this allele,
be the sum of the initial weights over all nodes who inherit allele $a$.
Note that if there has been only one mutation at the site,
then $\aw_j(a)$ is equal to the subtree weight of the node that the mutation producing $a$
appeared above, 
while more generally $\aw$ is not necessarily a subtree weight 
but is easily computable from them.
If the site has only one mutation, which produced allele $a$
at node $i$, then $\aw_j(a) = \nw(i)$.
Then, the statistic computed at site $j$ is
\begin{align}
    \site(f, \iw)_j
    &=
    \sum_{a \in D_j} \aw_j(a) ,
\end{align}
where the sum is over the \emph{derived} alleles $D_j$ at site $j$.
A summary statistic across $L$ sites is
\begin{align}
    \bar \site(f, \iw)
    &=
    \frac{1}{L} \sum_{j=1}^L \sum_a \aw_j(a) .
\end{align}
We call this the \textbf{site statistic} associated with $\iw$ and $f$.

\paragraph{Why the derived alleles?}
We have defined the statistic to sum over only derived alleles,
i.e., alleles not shared with the root of the tree.
At a biallelic site, this is equivalent to knowing the derived allele frequency
among any group of samples,
and such a statistic will be unchanged under recoding
if the summary function is unchanged after exchanging $p$ for $1-p$.
(XXX what is $p$)

@@

\plr{note LATER about what this does with triallelic sites}

@@

@@

\plr{Introduce branch stats.}

The \textbf{branch statistic} associated with the branch of length $t$
above node $u$ on a tree of length $\ell$,
with subtree weights $x$,
is
\begin{align}
    \branch(f, \iw)_u
    &=
    \ell t  f(\nw(u))  .
\end{align}
This is the summary value that would be added to a site statistic
if a single mutation occurred on this branch,
multiplied by the expected number of mutations 
(up to an overall constant related to mutation rate).

>>>>

Theory and computation of statistics of DNA sequences

@@

% summary
In this paper, we present a general framework for defining and efficiently computing
``single-site'' genetic statistics,
i.e., statistics of aligned genome sequence that can be expressed as averages over values computed
separately for each site.
Patterns of genetic polymorphism in a population
are determined by the population's history of Mendelian inheritance,
as summarized by the underlying genealogical trees,
and so it is natural (and helpful) to express these statistics
in terms of the structure of the trees themselves.
This perspective turns out to both suggest efficient algorithms
for computing these statistics on tree sequences,
and allows us to treat single-site genetic statistics in a more general framework
of summaries of tree shape and inheritance.
The methods are implemented in the \tskit{} library.

@@

Suppose we have genotyped a collection of $N$ (homologous) chromosomes
at a sequence of $L$ loci along the genome,
\jk{Maybe use $n$ here? We use it below}
and wish to compute statistic of the allele frequency spectrum.
In other words, if the frequency of the derived allele at a site $i$ is $p_i$,
and the derived allele frequency spectrum
-- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $f(p)$,
then for some function $h()$ we want to compute $\sum_p f(p) h(p)$.
Since this is a simple average over sites,
we can compute this, summing over the $L$ sites, as
$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L h(p_i)$.
(For the moment we ignore sites with more than two alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
occur below that mutation in the tree (and accounting for those that subsequently mutate).
What we do here is a generalization of this idea.

@@

Roughly speaking, we will 
(a) propagate ``weights'' additively up each tree,
(b) summarize these weights on each branch with a ``summary function'',
and (c) aggregate these summaries somehow
(e.g., averaged in windows).

@@

The basic ingredients that we use to summarize tree shape and genotype patterns
are the ``weights''.
For instance, we could calculate for a given tree how many of 
a certain set $S$ of samples are below each of its edges
by (a) assigning each of these samples weight 1, and other nodes weight zero, then
(b) finding the total weight of all nodes in the subtree below each edge.
This total weight is equal to the number of target samples below that edge,
and it can be calculated efficiently by propagating weights up the tree.
It turns out to be useful to allow more general weights.

\begin{definition}[Subtree weights]
    Given a list of \emph{node weights} $w$, and a tree $T$,
    define the \emph{subtree weight} $x_u$ for each node $u$ to be the sum of all node weights
    of every node below $u$ in the tree (including $u$ itself):
    \begin{align*}
        \nw(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
    \end{align*}
    where $v \le_T u$ if $u$ is on the path from $v$ back to the root of $T$.
\end{definition}

We will use weights of the form described above frequently enough
that we give them a special name --
for a set $S$ of samples,
the \emph{indicator weights of $S$} are the weights $\bone_S$ with
$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.

@@

The first method of summarizing uses subtree weights
to compute a summary for each variant site in the genome.
To compute the statistic at site $j$,
for each allele $a$ seen at this site,
let $\aw_j(a)$, the weight associated with this allele,
be the sum of the initial weights over all nodes who inherit allele $a$.
Note that if there has been only one mutation at the site,
then $\aw_j(a)$ is equal to the subtree weight of the node that the mutation producing $a$
appeared above, 
while more generally $\aw$ is not necessarily a subtree weight 
but is easily computable from them.
If the site has only one mutation, which produced allele $a$
at node $u$, then $\aw_j(a) = \nw(u)$.
Then, the statistic computed at site $j$ is
\begin{align}
    \site(f, \iw)_j
    &=
    \sum_{a \in D_j} \aw_j(a) ,
\end{align}
where the sum is over the \emph{derived} alleles $D_j$ at site $j$.
A summary statistic across $L$ sites is
\begin{align}
    \bar \site(f, \iw)
    &=
    \frac{1}{L} \sum_{j=1}^L \sum_a \aw_j(a) .
\end{align}
We call this the \textbf{site statistic} associated with $\iw$ and $f$.

\paragraph{Why the derived alleles?}
We have defined the statistic to sum over only derived alleles,
i.e., alleles not shared with the root of the tree.
This allows statistics in principle to differentiate ancestral from derived alleles,
information which is available in practice (albeit noisily).
If all sites are biallelic, and the sum of all weights is $\bar \iw = \sum_u \iw(u)$,
then any statistic defined for an $f$ with $f(x) = f(\bar \iw - x)$
will in fact not change if the identity of ancestral/derived alleles is flipped.

@@

@@

\paragraph{Polyallelic sites}
The statistics defined above map exactly onto those standard in population genetics
when all sites are biallelic. 
If there is a consensus in the field
for how to use sites with more than two alleles,
it is to reduce to biallelic data, by either discarding other sites
or by marking all non-ancestral alleles as (the same) ``derived'' allele.
We choose to define statistics in a way that still makes formal sense for polyallelic sites,
so that for instance a site with ancestral allele $A$ and derived alleles $C$ and $T$
would have site statistic $f(\aw(C)) + f(\aw(T))$.
This definition is simple, but we make no claims that it is the best or most natural one.

@@

Genetic variation is informative about many processes
precisely because it tells us about the underlying patterns of genealogical relatedness.
In other words, the genomes are useful in so far as they tell us about the trees.
In the case where we actually have the trees, or a good proxy for them,
it is natural to summarize them directly, rather than working indirectly with the genotypes.
If we assume that no two mutations occur at the same genomic position --
i.e., the \emph{infinite sites model} --
then there is a natural correspondence between summaries of genotypes and summaries of tree shape.
If mutations occur at a constant rate in time and along the genome,
then the \emph{expected} number of mutations that occur somewhere along a branch of a tree
over some segment of genome
is equal to the mutation rate multiplied by the length of the segment and by the length of the branch.
In other words, the ``area'' of a branch in a tree of the tree sequence,
defined as its span (right minus left endpoint) multiplied by its length (parent time minus child time)
is equal to the expected number of mutations that will land on it, up to the mutation rate.
If the mutation rate is constant,
then this gives us an isometry --
measuring tree distances in branch lengths versus in numbers of mutations
are equal in expectation, up to a constant.

With this in mind, we define the
\textbf{branch statistic} associated with the branch of length $t$
above node $u$ on a tree of length $\ell$,
with subtree weights $x$,
is
\begin{align}
    \branch(f, \iw)_u
    &=
    \ell t f(\nw(u))  .
\end{align}
This is the summary value that would be added to a site statistic
if a single mutation occurred on this branch,
multiplied by the expected number of mutations 
(up to an overall constant related to mutation rate).

A summary statistic across a region of the genome of length $L$ is
the average across this bit of genome WRITE THIS SOMEHOW
\begin{align}
    \bar \branch(f, \iw)
    &=
    \frac{1}{L} \sum_{T \in [0,L]} 
\end{align}
We call this the \textbf{site statistic} associated with $\iw$ and $f$.
